---
title: "RNAexpr Survival analysis BRCA"
output:
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, root.dir = "/home/deborah/Home")
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(tidyverse)
library(survival)
library(survminer)
#library(psych)
library(dplyr)
library(stringr)
library(data.table)
#library(tableone)
library(knitr)
library(kableExtra)
```


```{r echo=FALSE, warning=FALSE, message=FALSE}
################ KM curves
plot_surv=function(x,Df,subset,tit){
    t='end'
    my.formula<-as.formula(paste("Surv(time=",t,",event=",event,")~",x,sep=''))
    sd <- survdiff(my.formula, data = Df[subset,])
    p.val.KM<-1 - pchisq(sd$chisq, length(sd$n) - 1)
    p.val.KM<-formatC(p.val.KM, format = "e", digits = 2)
    surv_item= surv_fit(my.formula,data=Df[subset,])
    ggsurv<-ggsurvplot(surv_item, data=Df[subset,], main = "Survival curve",
               #fun = "cumhaz",
              ggtheme = theme_bw(),xlim=c(0.8*min(Df[,t],na.rm = TRUE),round(max(Df[,t],na.rm = TRUE))),
              legend.title = "CHD1L expr. levels: ",legend.labs = c("Mid", "Low","High"),legend = "right")
    ggsurv$plot <- ggsurv$plot+
      annotate("text",x=round(0.9*max(Df[,t],na.rm = TRUE)),y=0.7,label=paste("P.val=",p.val.KM,sep=''),size=4)+
      xlab("Days since diagnosis")+
      ggtitle(tit)
    ggsurv
    
}
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
setwd("/home/deborah/Documents/Crick/Projects/Coxph_RNA")
  #"/home/schneid/Documents/Crick/Projects/Coxph_RNA")

df_p.BRCA<-read.table(
  "BRCA_expression_survival_Mar2020.txt",
  sep="\t", header=TRUE)

#################################
receptors<-c('ER_status','PR_status','HER2_status')#,'her2_outcome')

rownames(df_p.BRCA)<-df_p.BRCA$PatientID
df_p.BRCA$STATUS<-as.factor(df_p.BRCA$STATUS)
df_p.BRCA<-df_p.BRCA[df_p.BRCA$TIME>0,]

#################################
#################### downloaded from TCGA 18/04/20
df_p.BRCA.or<-read.table(
   "BRCA_clinical_042020.tsv",
   sep="\t",fill=TRUE, header=TRUE)
r#################### downloaded from TCGA 18/04/20
################################
keep.3<-c("submitter_id","vital_status","days_tobirth","days_to_last_follow_up","days_to_death")
df_p.BRCA.or<-df_p.BRCA.or[,keep.3]
for(col in c("days_to_birth","days_to_last_follow_up","days_to_death")){
 df_p.BRCA.or[,col]<-as.numeric(levels(df_p.BRCA.or[,col])[df_p.BRCA.or[,col]])
}
df_p.BRCA.or<-df_p.BRCA.or[!duplicated(df_p.BRCA.or$submitter_id),]
#
df_p.BRCA.or<-df_p.BRCA.or[!is.na(df_p.BRCA.or$days_to_birth),]
df_p.BRCA.or$Age.yr<-(-1)*df_p.BRCA.or$days_to_birth/365
#
df_p.BRCA.or$end<-ifelse(df_p.BRCA.or$vital_status=="Alive",df_p.BRCA.or$days_to_last_follow_up,df_p.BRCA.or$days_to_death)
df_p.BRCA.or<-df_p.BRCA.or[(!is.na(df_p.BRCA.or$end))&(df_p.BRCA.or$end>0),]
df_p.BRCA.or$end.yr<-df_p.BRCA.or$end/365+df_p.BRCA.or$Age.y


######
all.genes<-fread(
  "BRCA_tumouronly.RSEM_genes_normalized_data.txt",
  sep="\t", select=1)

# ## find genes of interest
genes<-c('ATM','BRCA1','BRCA2','CHD1L')
c1<-lapply(genes, function(x) grep(x, all.genes$ID))
tmp<-(all.genes$ID[unlist(c1[1])]=='ATM')
c1[1]<-unlist(c1[1])[tmp]
c1=unlist(c1)

df_rsem.BRCA<-fread(
  "BRCA_tumouronly.RSEM_genes_normalized_data.txt",
  sep="\t",header = TRUE)

# #
df_rsem.BRCA=df_rsem.BRCA[c1,]
df_rsem.BRCA=as.data.frame(t(df_rsem.BRCA))
colnames(df_rsem.BRCA)<-genes
rownames(df_rsem.BRCA)<-str_replace_all(rownames(df_rsem.BRCA),"[.]", "-")
df_rsem.BRCA=df_rsem.BRCA[-1,]
indx <- sapply(df_rsem.BRCA, is.factor)
df_rsem.BRCA[indx] <- lapply(df_rsem.BRCA[indx], function(x) as.numeric(as.character(x)))
```

```{r echo=FALSE, warning=FALSE, message=FALSE}
rownames(df_rsem.BRCA)<-str_sub(rownames(df_rsem.BRCA),end=12)
Ids=intersect(rownames(df_p.BRCA),rownames(df_rsem.BRCA))
rownames(df_p.BRCA.or)<-df_p.BRCA.or$submitter_id
Ids.1<-intersect(rownames(df_p.BRCA.or),rownames(df_rsem.BRCA))

#### Merge 
Df.BRCA<-merge(df_p.BRCA.or,df_rsem.BRCA,by="row.names")
rownames(Df.BRCA)<-Df.BRCA$Row.names
rownames(df_p.BRCA)<-df_p.BRCA$PatientID
Df.BRCA<-merge(Df.BRCA,df_p.BRCA[,receptors],by="row.names")


### Fix RNA expr vars.
for(g in genes){
  Df.BRCA[,g]<-as.numeric(as.character(Df.BRCA[,g]))
}

#### create cat. of RNA expr. 
for(var in genes[-1]){
  Df.BRCA$tmp=cut(Df.BRCA[,var],breaks=quantile(Df.BRCA[,var],c(0,0.25,0.75,1))) # split around quarters 
  levels(Df.BRCA$tmp)<-c('low','mid','high')
  Df.BRCA$tmp<-relevel(Df.BRCA$tmp, ref='mid')
  new_var<-paste(var,'_cat',sep='')
  ncols=length(colnames(Df.BRCA))
  colnames(Df.BRCA)[ncols]<-new_var
}
# split ATM around median
Df.BRCA$ATM_cat<-cut(Df.BRCA$ATM,breaks=quantile(Df.BRCA$ATM,c(0,0.5,1)))
levels(Df.BRCA$ATM_cat)<-c('low','high')
Df.BRCA$ATM_cat<-relevel(Df.BRCA$ATM_cat, ref='low')
#
Df.BRCA$HRD<-ifelse((Df.BRCA$BRCA1_cat=='low')|(Df.BRCA$BRCA2_cat=='low' ),1,0)
Df.BRCA<-Df.BRCA[!is.na(Df.BRCA$HRD),]
#
Df.BRCA$over65<-ifelse(Df.BRCA$Age.yr>=65,1,0)

```

```{r echo=FALSE, warning=FALSE, message=FALSE}

levels(Df.BRCA$vital_status)<-c(0,1) #0 alive -no event - 1 dead - event

##### surv items set here #####
event="vital_status"
t="Age.yr,time2=end.yr"

surv.gen<-paste("Surv(time=",t,",event=",event,")~",sep='')
```

## Cohort description
Cohort description

```{r echo=FALSE, warning=FALSE, message=FALSE}
genes_cat=lapply(genes, function(x) paste(x,"_cat",sep=''))

Df.table<-Df.BRCA
Df.table<-Df.table[,colnames(Df.table) != "Row.names"]
Df.table<-Df.table %>% 
  rename( Dead=vital_status, Censoring_Time=end, Censoring_age=end.yr)
varstofact<-unlist(c("Dead","over65","HRD",receptors))
Df.table[varstofact] <- lapply(Df.table[varstofact], factor)
Vars<-c(varstofact,"Age.yr","Censoring_Time","Censoring_age",genes)
tableOne <- CreateTableOne(vars = Vars,data=Df.table)

T1<-kableone(tableOne)

T1 %>%
  kable_styling(bootstrap_options = "striped", full_width = F)

#### HRD status: low expr of BRCA 1 or 2 
```

Categories of gene expression:

* CHD1L, BRCA1 and BRCA2: low = bottom 25% of data sample. Mid: median 50% of sample. High: top 25% of sample
* ATM: low= bottom 50%; High= top 50%. Median = `r median(Df.BRCA$ATM) `

In subsequent regression analyses, the mid category for CHD1L, BRCA1 and BRCA2 is used as the reference category. For ATM gene expression, the low category is used as reference

HRD status is defined as having a low expression profile for BRCA1 AND/OR BRCA2 genes (bottom 25% of sample for either or both genes)

```{r echo=FALSE, warning=FALSE, message=FALSE}
qs<-lapply(Df.BRCA[,genes],function(x) quantile(x,c(0,0.25,0.75,1)))
d.qs <- data.frame(matrix(unlist(qs), nrow=length(qs), byrow=T))
colnames(d.qs)<-c('min','bottom 25%','top 25%','max')
rownames(d.qs)<-genes

signif(d.qs,3) %>%
   kable() %>%
   kable_styling()
  
```
### Exclusion criteria

Patients were removed from the analysis if they had any of the following:

* missing date of birth
* missing survival information (no death date + no last follow-up date + no vital status information)

## Logit regression on HRD status - association with CHD1L gene expression

```{r echo=FALSE, warning=FALSE, message=FALSE}
stdev=sd(Df.BRCA$CHD1L)
pretty_pres<-function(x,stdev){ 
  x <- summary(x)
  p.value<-signif(x$coefficients[,4], digits=2)
  wald.test<-signif(x$coefficients[,3], digits=2)
  beta<-signif(x$coefficients[,1], digits=3);#coeficient beta
  HR <- signif(exp(beta*stdev),4)
  res<-as.data.frame(cbind(HR, p.value))
  colnames(res)<-c("OR", 
                "p.value")
  return(res)
}

logit.model<-"HRD~CHD1L+Age.yr"
logit.res<-glm(as.formula(logit.model),data=Df.BRCA,family = 'binomial')
#
pretty_pres(logit.res,stdev) %>%
     kable(caption = "Logistic regression: HRD status vs CHD1L expr.") %>%
   kable_styling()

#
Df.plot<-Df.BRCA
Df.plot$HRD<-as.factor(Df.plot$HRD)
#
ttest.hrd<-t.test(CHD1L~HRD,data=Df.plot)
pp<-formatC(ttest.hrd$p.value, format = "e", digits = 2)
#
gg1<-ggplot(Df.plot[,-2],aes(x=HRD,y=CHD1L,fill=HRD))+
  geom_boxplot()+
  ylim(0,2000)+
  #geom_smooth(method='glm', formula= y~x)+
  annotate(geom="text",x=2, y=1, label=paste("P.val=",pp,sep=''))+
  annotate(geom="text",x=1, y=1500, label=paste("Mean=",signif(ttest.hrd$estimate[1],4),sep=''))+
  annotate(geom="text",x=2, y=1500, label=paste("Mean=",signif(ttest.hrd$estimate[2],4),sep=''))
gg1

```

### Association between ATM x CHD1L

```{r echo=FALSE, warning=FALSE, message=FALSE}
#
lm1<-summary(lm('CHD1L~ATM',data=Df.plot))
pp2<-lm1$coefficients[2,c(1,4)]

#
gg2<-ggplot(Df.plot[,-2],aes(x=ATM,y=CHD1L))+
  geom_point()+
  #ylim(0,2000)+
  geom_smooth(method='glm', formula= y~x)+
  annotate(geom="text",x=5000, y=2500, label=paste("Cor=",signif(pp2[1],3),', P.val=',signif(pp2[2],4),sep=''))
gg2
```

### Association  between BRCA2 x CHD1L

```{r echo=FALSE, warning=FALSE, message=FALSE}
lm2<-summary(lm('CHD1L~BRCA1*BRCA2',data=Df.plot))
pp3<-lm2$coefficients[c(2,3),c(1,4)]
corr<-formatC(pp3[,1],digits=3)
pp3[,2]<-formatC(pp3[,2], format = "e", digits = 2)
#
gg2<-ggplot(Df.plot[,-2],aes(x=BRCA2,y=CHD1L))+
  geom_point()+
  #ylim(0,2000)+
  geom_smooth(method='lm', formula= y~x)+
  annotate(geom="text",x=500, y=7500, label=paste("Cor=",corr[2],', P.val=',pp3[2,2],sep=''))
gg2
#BRCA2
lm.brca2<-summary(lm('CHD1L~BRCA2_cat',data=Df.plot))
pp4<-lm.brca2$coefficients[,4]
pp4<-formatC(pp4, format = "e", digits = 2)
ggplot(Df.plot[,-2],aes(x=BRCA2_cat,y=CHD1L,fill=BRCA2_cat))+
  geom_boxplot()+
  ylim(0,2000)+
  annotate(geom="text",x=2, y=0, label=paste('P.val=',pp4[2],sep=''))+
  annotate(geom="text",x=3, y=0, label=paste('P.val=',pp4[3],sep=''))
```

### Association  between BRCA1 x CHD1L

```{r echo=FALSE, warning=FALSE, message=FALSE}
Df.plot<-Df.plot[!is.na(Df.plot$BRCA1_cat),]
gg2<-ggplot(Df.plot[,-2],aes(x=BRCA1,y=CHD1L))+
  geom_point()+
  #ylim(0,2000)+
  geom_smooth(method='lm', formula= y~x)+
  annotate(geom="text",x=500, y=7500, label=paste("Cor=",corr[1],', P.val=',pp3[1,2],sep=''))
gg2
#BRCA2
lm.brca2<-summary(lm('CHD1L~BRCA1_cat',data=Df.plot))
pp4<-lm.brca2$coefficients[,4]
pp4<-formatC(pp4, format = "e", digits = 2)
ggplot(Df.plot[,-2],aes(x=BRCA1_cat,y=CHD1L,fill=BRCA1_cat))+
  geom_boxplot()+
  ylim(0,2000)+
  annotate(geom="text",x=2, y=0, label=paste('P.val=',pp4[2],sep=''))+
  annotate(geom="text",x=3, y=0, label=paste('P.val=',pp4[3],sep=''))
```

## Cox survival analyses

```{r echo=FALSE, warning=FALSE, message=FALSE}
#
genes_cat=lapply(genes, function(x) paste(x,"_cat",sep=''))
Df.model<-Df.BRCA[,unlist(c("vital_status","Age.yr","end","end.yr",'HRD',
                            'over65',genes,genes_cat,receptors))]
levels(Df.model$vital_status)<-c(0,1)
Df.model$vital_status<-as.numeric(Df.model$vital_status)-1
Df.model$HRD<-as.factor(Df.model$HRD)

pretty_pres_coxph<-function(x){ 
  x <- summary(x)
  p.value<-signif(x$coefficients[,5], digits=2)
  wald.test<-signif(x$coefficients[,4], digits=2)
  beta<-signif(x$coefficients[,1], digits=3);#coeficient beta
  HR <-signif(x$coefficients[,2], digits=3);#exp(beta)
  HR.confint.lower <- signif(x$conf.int[,"lower .95"],3) #signif( ,2)
  HR.confint.upper <- signif(x$conf.int[,"upper .95"],3)
  HR <- paste0(HR, " (", 
               HR.confint.lower, "-", HR.confint.upper, ")")
  res<-as.data.frame(cbind(HR, p.value))
  colnames(res)<-c("HR (95% CI for HR)", 
                "p.value")
  return(res)
}

split_cat_plot<-function(x,names_2){
  if(x=='mid'){
    keep<-unlist(lapply(names_2, function(x) str_count(x,"-")))
    keep<-which(keep==1)
    names.low<-names_2[keep]
    names.low<-setdiff(names.low,grep("CHD1L",names.low,value=TRUE))
    df<-res[res$Variables %in% names.low,]
    gene<-unlist(str_split(names.low[1],'-'))[1]
    names.low.2<-c(names.low,paste(gene,'-mid',sep=''))
    cat="CHD1L-mid"
  }else{
    cat<-paste("CHD1L-",x,sep='')
    names.low<-grep(cat,names_2,value = TRUE)
    names.low.2<-gsub(paste("(",cat,")[a-zA-Z1-2-]",sep=''),'',names.low)[-1]
    gene<-unlist(str_split(names.low.2[1],'-'))[1]
    names.low.2<-c(paste(gene,'-mid',sep=''),names.low.2)
    df<-res[res$Variables %in% names.low,]
  }
  #
  df$Variables<-droplevels(df$Variables)
  levels(df$Variables)<-names.low.2
  # if(x=='mid'){
  #   df<-rbind(df,c(names.low.2[3],1,1,1))
  # }
  G<-ggplot(df,aes(x=Variables,y=HR))+geom_point(colour = "red", size = 3)+
    geom_point(aes(x=Variables,y=LL))+
    geom_point(aes(x=Variables,y=UL))+
    geom_hline(yintercept = 1)+
    #theme(axis.text.x = element_text(angle = 90))+
    xlab('')+ggtitle(cat)
  return(G)
}

pretty_plot_coxph<-function(x){ 
  x <- summary(x)
  p.value<-x$coefficients[,5]
  res<-data.frame(Variables=rownames(x$coefficients))
  res$HR <-as.numeric(x$coefficients[,2])
  res$LL <- as.numeric(x$conf.int[,"lower .95"])
  res$UL <- as.numeric(x$conf.int[,"upper .95"])
  # res$p.value<-ifelse(p.value<0.001,'***',
  #                 ifelse(p.value<0.01,'**',
  #                        ifelse(p.value<0.05,'*','')))
  #res<-melt(res)
  names<-levels(res$Variables)
  names_2<-str_split(names,"_cat|:")
  names_2<-sapply(names_2,function(x) paste(x, collapse='-' ))
  levels(res$Variables)<-names_2
  #
  gg.low<-split_cat_plot('low',names_2)
  gg.high<-split_cat_plot('high',names_2)
  gg.mid<-split_cat_plot('mid',names_2)
  Gg<-ggarrange(gg.low,gg.mid,gg.high,ncol=3)
  return(Gg)
}
#
```

### Cox survival analysis for BRCA Cancer - ATM x CHD1L

```{r echo=FALSE, warning=FALSE, message=FALSE}
cox.0.atm<-coxph(as.formula(paste(surv.gen,' CHD1L_cat+ATM_cat',sep='')), data = Df.model) 

cox.1.atm<-coxph(as.formula(paste(surv.gen,' CHD1L_cat*ATM_cat',sep='')), data = Df.model) 
#ggforest(cox.1.atm)

####
res.cox.atm <- coxph(as.formula(paste(surv.gen,' CHD1L_cat*ATM_cat+HRD',sep='')), data = Df.model)

res.atm<-anova(cox.0.atm,cox.1.atm)#,res.cox.atm)
res.atm
p.inter<-res.atm$`P(>|Chi|)`[2]
tmp<-which(res.atm$`P(>|Chi|)`<0.05)
if(length(tmp)==0){
  ggforest(cox.0.atm)
}else{
 pretty_plot_coxph(cox.1.atm)
}

```

P-value for effect of CHD1L X ATM on survival: `r formatC(p.inter, format = "e", digits = 2)`

### Cox survival analysis for BRCA Cancer - HRD x CHD1:

```{r echo=FALSE, warning=FALSE, message=FALSE}
cox.0.hrd<-coxph(as.formula(paste(surv.gen,' CHD1L_cat+HRD',sep='')), data = Df.model) 

cox.1.hrd<-coxph(as.formula(paste(surv.gen,' CHD1L_cat*HRD',sep='')), data = Df.model) 

res.cox.hrd <- coxph(as.formula(paste(surv.gen,' CHD1L_cat*HRD+ATM_cat',sep='')), data = Df.model) #=start,time2=

res.hrd<-anova(cox.0.hrd,cox.1.hrd)#,res.cox.hrd)
res.hrd
p.inter<-res.hrd$`P(>|Chi|)`[2]
tmp<-which(res.hrd$`P(>|Chi|)`<0.05)
if(length(tmp)==0){
  ggforest(cox.0.hrd)
}else{
 pretty_plot_coxph(cox.1.hrd)
}
```

P-value for effect of CHD1L X HRD on survival: `r formatC(p.inter, format = "e", digits = 2)`

### Cox survival analysis for BRCA Cancer - BRCA2 x CHD1L

```{r echo=FALSE, warning=FALSE, message=FALSE}
cox.0.brca2<-coxph(as.formula(paste(surv.gen,' CHD1L_cat+BRCA2_cat',sep='')), data = Df.model) 

cox.1.brca2<-coxph(as.formula(paste(surv.gen,' CHD1L_cat*BRCA2_cat',sep='')), data = Df.model) 

res.brca2<-anova(cox.0.brca2,cox.1.brca2)
res.brca2
p.inter<-res.brca2$`P(>|Chi|)`[2]
tmp<-which(res.brca2$`P(>|Chi|)`<0.05)
if(length(tmp)==0){
  ggforest(cox.0.brca2)
# pretty_pres_coxph(res.cox) %>%
#   kable() %>%
#   kable_styling()
}else{
  pretty_plot_coxph(cox.1.brca2)
  setEPS()
  postscript("cox_BRCA2xACL1.eps",width = 10, height = 7)
  pretty_plot_coxph(cox.1.brca2)
  dev.off()
}
```

P-value for effect of CHD1L X BRCA2 on survival: `r formatC(p.inter, format = "e", digits = 2)`

### Cox survival analysis for BRCA Cancer - BRCA1 x CHD1L

```{r echo=FALSE, warning=FALSE, message=FALSE}
cox.0.brca2<-coxph(as.formula(paste(surv.gen,' CHD1L_cat+BRCA1_cat',sep='')), data = Df.model) 

cox.1.brca2<-coxph(as.formula(paste(surv.gen,' CHD1L_cat*BRCA1_cat',sep='')), data = Df.model) 

res.brca2<-anova(cox.0.brca2,cox.1.brca2)
res.brca2
p.inter<-res.brca2$`P(>|Chi|)`[2]
tmp<-which(res.brca2$`P(>|Chi|)`<0.05)
if(length(tmp)==0){
  ggforest(cox.0.brca2)
}else{
 pretty_plot_coxph(cox.1.brca2)
}
```

P-value for effect of CHD1L X BRCA1 on survival: `r formatC(p.inter, format = "e", digits = 2)`

## KM curves for all genes of interest + HRD status

```{r echo=FALSE, warning=FALSE, message=FALSE}
lapply(genes_cat, function(x) plot_surv(x,Df.model))
plot_surv('HRD',Df.model)
```

### KM curves for genes interaction
```{r echo=FALSE, warning=FALSE, message=FALSE, fig.width=12}
Df.model$BRCA2_mid<-cut(Df.model$BRCA2,breaks=quantile(Df.model$BRCA2,c(0,0.5,1)))
levels(Df.model$BRCA2_mid)<-c('low','high')
Df.model$BRCA1_mid<-cut(Df.model$BRCA2,breaks=quantile(Df.model$BRCA1,c(0,0.5,1)))
levels(Df.model$BRCA1_mid)<-c('low','high')
Df.model$ATM_mid<-cut(Df.model$ATM,breaks=quantile(Df.model$ATM,c(0,0.5,1)))
levels(Df.model$ATM_mid)<-c('low','high')
  setEPS()
  postscript("KM_BRCA2xACL1.eps",width = 10, height = 7)
plot_surv('interaction(CHD1L_cat,BRCA2_cat)',Df.model,(Df.model$BRCA2_cat=='low'),"Low BRCA2 expr - bottom 33% of cohort")
  dev.off()
  setEPS()
  postscript("KM_BRCA1xACL1.eps",width = 10, height = 7)
plot_surv('interaction(CHD1L_cat,BRCA1_cat)',Df.model,(Df.model$BRCA1_cat=='low'),"Low BRCA1 expr - bottom 33% of cohort")
    dev.off()
setEPS()
  postscript("KM_ATMxACL1.eps",width = 10, height = 7)
plot_surv('interaction(CHD1L_cat,ATM_cat)',Df.model,(Df.model$ATM_cat=='low'),"Low ATM expr - bottom 33% of cohort")
  dev.off()
```
